---
title: "Redwine Analysis"
author: "Tu Duong"
date: "Feb 19, 2018"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    toc_float: true
---


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape2)
library (GGally)
#library(gridExtra)
library(memisc)
library(ggcorrplot)
```

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data

setwd('/Users/tduong/Documents/R/eda-course-materials/EDA Project')
redwine=read.csv('wineQualityReds.csv')

```


This project explores a dataset of 1600 red wine samples 
to understand what physicochemical properties influence the wine's quality.

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plot1}

str(redwine)

```

There are 1599 wine samples. The first variable  is the sample index, then there 
are 11 variables describe wine characteristics, and a variable shows experts'
opion on the quality of the wine

```{r echo=FALSE, warning=FALSE}

summary(redwine)

```

This summary shows the range and the median of each characterics. It is
interesting to see the median quality is 6 the the range from 3 to 8.
There is no really bad wine (quality index 0-2) or really good wine (quality 
index 9-10)


```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=quality),data=redwine)+geom_histogram(binwidth=1)

```


```{r echo=FALSE, warning=FALSE}

table(redwine$quality)

```

Most of the wine in the sample have average quality (score 5-6)


```{r echo=FALSE, warning=FALSE}

#prepare to create a matrix of histogram, one for each variable
redwine.long<-melt(subset(redwine,select=names(redwine)!='X'))

ggplot(data = redwine.long, mapping = aes(x = value)) + 
    geom_histogram(bins = 10) + facet_wrap(~variable, scales = 'free_x')

```


With the histogram for each variable, we can quickly see the distributions.

Most variables are bell-shaped or left skewed.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(fixed.acidity),data=redwine)+geom_histogram(binwidth = 1)

```

Most samples have fixed acidity from 6-9 gram/liter, the the shape is slighty skewed 
to the left. There are some outlier with acidity higher than 13 gram/liter.


```{r, echo=FALSE, warning=FALSE}

ggplot(aes(volatile.acidity),data=redwine)+geom_histogram(binwidth = .1)

```

Volatile acidity has a bell-shape distribution, with outlier samples with acid
level higher than 1 gram/liter.



```{r, echo= FALSE, warning=FALSE,}

ggplot(aes(citric.acid),data=redwine)+geom_histogram(binwidth = .05)

```

Using binwidth 0.1 is does not give us a clear view on distribution. Using 
binwidth 0.01 reveals there are many sample with zero citric acid. 
It is quite rare to have citric acid level higher than 0.75g/l.



```{r, echo=FALSE, warning=FALSE}

ggplot(aes(residual.sugar),data=redwine)+geom_histogram(binwidth = .1)

```

Most samples have residual sugar from 2-3grams/liter. Interesting there is no 
sweet wine (sugar > 45gr/liter)


```{r, echo=FALSE, warning=FALSE}

ggplot(aes(residual.sugar),data=redwine)+geom_histogram(binwidth = .1)+
  xlim(c(0,4))

```

Removing the outlier with sugar higher than 2g/l, it is easier to see the bell curve 
distribution around 2g/liter.

```{r, echo=FALSE, warning=FALSE}

ggplot(aes(chlorides),data=redwine)+geom_histogram(binwidth = .01)

```

Chlorides has similar distribution to residual sugar, most samples has chlorides
level less than 0.1g/l. Some outlier samples have chloride up to 0.6g/l. I am wondreing
if those outliers has negative or positive influence to the quality of the wine,
if it has any influence at all?

```{r, echo=FALSE, warning=FALSE}

ggplot(aes(chlorides),data=redwine)+geom_histogram(binwidth = .01)+
  xlim(c(0,0.15))

```

A zoom in to the chlorides level of the majority of the sample fluctuate around
0.75g/l.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(free.sulfur.dioxide),data=redwine)+geom_histogram(binwidth = 1)

```

Free surfur dioxide is mainly below 20mg/l.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(total.sulfur.dioxide),data=redwine)+geom_histogram(binwidth = 10)

```

Total sulfur dioxide is left skewed. Most samples have total silfur dioxide below
150mg/l.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(total.sulfur.dioxide),data=redwine)+
  geom_histogram()+
  scale_x_log10()

```


Transform total sulfur dioxide to log scale shows a close to normal distribution.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(density),data=redwine)+geom_histogram(binwidth = .001)

```

Density follows bell curved distribution with a median of 0.9968g/cm^3.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(pH),data=redwine)+geom_histogram(binwidth = .1)

```

pH distribution follow bell curve with median of 3.310.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(sulphates),data=redwine)+geom_histogram(binwidth = 0.1)

```



```{r echo=FALSE, warning=FALSE}

ggplot(aes(sulphates),data=redwine)+geom_histogram(binwidth = 0.05)+
  scale_x_log10(breaks=c(0.45,0.55,0.62,0.73,0.83))

```

After transfrom x axix to log10 scale, the distriubtion is bell curved. This
suggests a multiplicative effect for this variable.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(alcohol),data=redwine)+geom_histogram(binwidth = 0.5)

```


# Univariate Analysis

### What is the structure of your dataset?

The dataset contains about 1600 samples, each has 12 variables. All features
are expressed in numbers (quality in integer, which is a categorical variable).

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is quality. We would like to know if and how much
the physicochemical properties can be used to predict the quality of the wine.

The median quality is 6, wine sample quality range from 3to 8.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

All other features apart from X seems to be independent from each other and 
they all should be used to investigate how they influence quality.

### Did you create any new variables from existing variables in the dataset?

No new variables are created as all influencing features seems independent 
from each other.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Dataset is clean so data cleaning is not required. I log transfromed the left
skew variables to check if they have a clear multiplicative type.


# Bivariate Plots Section

Correlation matrix

```{r echo=FALSE, warning=TRUE}
redwine.subset <- subset(redwine, select = -c(X))
corr <- round(cor(redwine.subset), 1)
ggcorrplot(corr,hc.order = TRUE, outline.col = "white",lab=TRUE)
```

```{r echo=FALSE, warning=TRUE}
round(cor(redwine.subset), 3)
```

I prepared the correlation matrix to see the relationships among various 
variables before selecting the important ones to continue the investigation. 
In particular I'd like to see how physicochemical variables influence quality.
I am surprised to see there is no strong colleration between quality versus 
the input variables. The highest correlations are with alcohol level (0.476), 
volatile acidity (-0.391), sulfates (0.251) and citric acid (0.226).

Although I can understand high level of volatile acidity makes the wine sour
and therefore unpleasant, I never thought that alcohol level have high 
colleration to quality.

Now I start to wonder about the factors that can influence the wine quality 
that are not captured by the variables in the dataset.For example, the grape 
type, grape ripping level, age of the vine, or the wine region could create 
some charactrs that we do not capture here (I am thinking about Bordeaux and 
Napa Valley here). 

However, there are interesting relationships among the physicochemical variables.
For example, fixed acidity collerate strongly with density (0.668) and pH
(-0.683), and citric acid (0.672).

I also wondered how the outliers in the physicochemical properties impact wine
quality? Will they create exceptional good wine or bad wine?

My next analysis will go deep into these interesting relationships.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality),data=redwine)+geom_point()

```

Overplotting problem to be fixed.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality),data=redwine)+
  geom_jitter(alpha=0.2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))
  #geom_smooth(method='lm')

```

Using jitter shows an interseting insight: most of the wine with alcohol
below 10% are of average quality. Higer quality wine (7 and above) have 
acohol level of 10% or more.


```{r echo=FALSE, warning=FALSE}

redwine$quality.category=as.factor(redwine$quality)
ggplot(aes(x=quality.category,y=alcohol),data=redwine)+geom_boxplot()+
  coord_flip()

```

This box plot makes the relationship look clearer. The median alcohol levels
of wine with quality index 7 and 8 are higher than those with lower quality.

However, there are many average quality sample with high alcohol content. What
are other variables that can explain this?

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=volatile.acidity,y=quality),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$volatile.acidity,0.01),quantile(redwine$volatile.acidity,0.99)))

```

Low correlation between the two variables.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=quality.category,y=volatile.acidity),data=redwine)+
  geom_boxplot()+
  coord_flip()

```


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=sulphates,y=quality),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$sulphates,0.01),quantile(redwine$sulphates,0.99)))

```


```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=quality.category,y=sulphates),data=redwine)+
  geom_boxplot()+
  coord_flip()

```

Sulphates level has a lower correlation with quality. In general, the higher 
sulphates, the higher the wine quality - up to a point. No good wine has 
sulphates level above 1.5g/l.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=citric.acid,y=quality),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$citric.acid,0.01),quantile(redwine$citric.acid,0.99)))

```


```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=quality.category,y=citric.acid),data=redwine)+
  geom_boxplot()+
  coord_flip()

```

Similar to sulphates, citric acid has a weak correlation to quality, but overall
higher quality wines have higher median citric acid.

Now I look into relationships among variables other than quality. 
There are interesting relationship among the physicochemical variables.
For example, fixed acidity collerate strongly with density (0.668), pH
(-0.683), and citric acid (0.672).

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=fixed.acidity,y=density),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$fixed.acidity,0.01),quantile(redwine$fixed.acidity,0.99)))+
  geom_smooth(method='lm')

```



```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=fixed.acidity,y=pH),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$fixed.acidity,0.01),quantile(redwine$fixed.acidity,0.99)))+
  ylim(c(quantile(redwine$pH,0.01),quantile(redwine$pH,0.99)))+
  geom_smooth(method='lm')

```

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=fixed.acidity,y=citric.acid),data=redwine)+geom_jitter(alpha=0.2)+
  xlim(c(quantile(redwine$fixed.acidity,0.01),quantile(redwine$fixed.acidity,0.99)))+
  geom_smooth(method='lm')

```

There is a correlation, but there are varied level of fixed acidity with 
citric acid level of 0g/l, 

Now I want to understand how the outliers impact quality. I don't like sweet
wine, so I am checking for the sugar level outlier.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=residual.sugar,y=quality),data=redwine)+geom_jitter(alpha=0.2)

#+ xlim(c(quantile(redwine$citric.acid,0.01),quantile(redwine$citric.acid,0.99)))

```

This plot indicate that good wine does not have residual sugar higher than 8g/l.


```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=chlorides,y=quality),data=redwine)+geom_jitter(alpha=0.2)

```

Same as residual sugar, exceptionally high chlorides does not make a good wine.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There are weak correlation between quality and other physicochemical qualities.
It seems that there are many factors could influence the quality of the wine, 
and perhaps the factors included in the dataset is not enough to predict the 
wine's quality.

Having said that, there are some interesting relationships between wine quality 
and the physicochemical features. The strongest correlation (0.476) is with
alcohol level, where the dataset suggests that high quality wine (7/8) have 
median alcohol level higher than 11% while low quality wine (3-5) have median 
alcohol content less than 11%.

On the other hand, the is a negative correlation between quality and volatile 
acidity. The box plot shows the higher the acidity, the lower the wine quality.

It is also interesting to note the sweet spot for sulphates. Good wines have
sulphates level around 0.7-0.8g/l. Sulphates level outside this range seems to 
to result in bad or average wines. Similarly, the sweet spot for citric acid is
around 0.3 -0.5 gram/liter.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There are interesting relationship among the physicochemical variables.
For example, fixed acidity collerate strongly with density (0.668), pH
(-0.683), and citric acid (0.672) .

### What was the strongest relationship you found?
Quality of wine is positively, although not strongly, correlated with alcohol 
content. 

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality, color=volatile.acidity),data=redwine)+
  geom_jitter(alpha=0.2,size=2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))

```

The plot shows there is a week relationship between alcohol and volatile acidity.

```{r}

ggplot(aes(x=volatile.acidity, y=alcohol),data=redwine) + 
  geom_jitter(alpha=0.2)+
  facet_wrap( ~ quality,ncol=3)

```

This plot shows wine with quality index 8 has higher alcohol and low volatile 
acidity.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality, color=sulphates),data=redwine)+
  geom_jitter(alpha=0.2,size=2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))

```

It appears that high wine quality have high sulphates.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality, color=residual.sugar),data=redwine)+
  geom_jitter(alpha=0.5,size=2)+
  xlim(quantile(redwine$alcohol,0.05),quantile(redwine$alcohol,0.95))

```



```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality, color=density),data=redwine)+
  geom_jitter(alpha=0.2,size=2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))

```


```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=quality, color=fixed.acidity),data=redwine)+
  geom_jitter(alpha=0.2,size=2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))


```

The above charts were not able to reveal any significant relationships among
the variables.

```{r echo=FALSE, warning=FALSE}

ggplot(aes(x=fixed.acidity,y=density, color=citric.acid),data=redwine)+
  geom_point(alpha=0.2)+
  xlim(quantile(redwine$fixed.acidity,0.01),quantile(redwine$fixed.acidity,0.99))+
  geom_smooth(method='lm')

```

This chart is more interesting as it shows (i) there is a positive correlation between
fixed acidity and density. and (ii) for a given level of fixed acidity, the higher level of
citric acid, the lower the density.


Now I want to quantify the relationship of various input variables with quality.
I will convert to convert the factor quality into a numerical variable and run
a linear regession model. The assumption that the final quality score is rounded
to the nearest interger number.


```{r echo=FALSE, warning=FALSE}

m1<-lm(quality~alcohol,data=redwine)
m2<-update(m1,~.+volatile.acidity)
m3<-update(m2,~.+sulphates)
m4<-update(m3,~.+citric.acid)
mtable(m1,m2,m3,m4)  
```

The model confirms that the most important variables (alcohol, volatile acidity
sulphates, and citric acid) can only explain 33.6% of quality. There is no 
additional improvement as the last variable citric acid did not improve  
R-square.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As suspected, due to the week correlation among variables, our multivariate
analysis only gain modest additional insights.


### Were there any interesting or surprising interactions between features?

While quality does not have strong correlation with other variables, fixed 
acidity has fairly strong correlation with density, pH, and citric acid.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

As commented previously, it seems quality is impacted by various factors, and 
some of those might not be included in the data set. Therefore, I created a 
linear regression model to access how much the top 4 physicochemical propereties 
explain the wine quality. The model is only able to explain one third of the 
variation in quality.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}

ggplot(aes(x=quality.category,y=alcohol, fill=I('green')),data=redwine)+
  geom_boxplot()+
  coord_flip()+
  ylab('Alcohol %')+
  xlab('Wine quality index')+
  ggtitle('Wine quality by alcohol content', subtitle = NULL)+
  ylim(c(9.0,14.0))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())



```

### Description One

It is surprising to see there is a relationship between wine quality and its 
alcohol content. Median alcohol content of higer quality wine (7 and 8) is
higher than 11%, while most lower wine quality have alcohol below 11%.

### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_Two}

ggplot(aes(x=alcohol,y=quality, color=I('blue')),data=redwine)+
  geom_jitter(alpha=0.2)+
  xlim(quantile(redwine$alcohol,0.01),quantile(redwine$alcohol,0.99))+
  xlab('Alcohol %')+
  ylab('Wine quality index')+
  ggtitle('Relationship between quality and alcohol', subtitle = NULL)+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```

### Description Two
Among the variables, alcohol content has highest correlation to wine quality.
However, the plot indicated that the relationship is weak. The top 4 variables
(alcohol, variable acidity, sulphates, and citric acid) can only explain 
one-third of the quality of redwine. THis indicates that there are many other
factors that influence the wine quality that is missing from the dataset (for 
example, grape types or regions)

### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x=fixed.acidity,y=density, color=citric.acid),data=redwine)+
  geom_point(alpha=0.2,size=2)+
  xlim(quantile(redwine$fixed.acidity,0.01),quantile(redwine$fixed.acidity,0.99))+
  geom_smooth(method='lm',color=I('black'))+
  xlab('Fixed Acidity (g/l)')+
  ylab('Density g/cm^3')+
  ggtitle('Density by fixed acidity and citric acid', subtitle = NULL)+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(color = "Citric acid (g/l)")

```

### Description Three

There is a strong positive correlation between fixed acidity and density
and for a given level of fixed acidity, the higher level of
citric acid, the lower the density.

------

# Reflection

The wine datasetcontains information on 1600 samples with 12 variable.
I started by understanding the individual variables in the data set, 
and then I created many plots to understand the relationship among variables, 
most important to find out the important variables  that might influence quality.

It is interesting to understand the values and distributions of the variables. 
The frustration startedafter I built the scatterplot matrix that shows weak 
correlations between quality and other variables. The multivariate analysis further 
confirm the hypothesis that there are might be so many factors influence wine 
quality, or the important factors (grape types, wine region) were not captured 
in the dataset. It is interesting to understant the relationship between alcohol
level with with quality, as well as volatile acidity with quality. 
Also, it is suprising to discover that there is a high correlation among fixed 
acidity, citric acid, and density.

For future work related to this dataset, I would like to include additional 
variables that could be important in determining the quality of the wine.
From my experience, the wine quality could depend on the grape variety (for 
example, merlot,cabernet sauvignon, shiraz, etc.), the wine region, and the year
of the wine. Adding these additional variable can help us to build a better 
predictive model of wine quality.



